\pkg{sdmTMB}: An \proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields

### 2023-09-20 doc/paper-jss/sdmTMB-paper.Rnw
Wrapping; complex mesh; new fig 1 Needs my local sdmTMB for mesh!

<<main-setup, include=FALSE, cache=FALSE>>=
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  cache.comments = TRUE,
  dev = "pdf",
  dpi = 140,
  optipng = "-strip all",
  highlight = FALSE
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
opts_chunk$set(prompt = TRUE)
options(replace.assign = TRUE, width = 72, prompt = "R> ")
knitr::render_sweave()
@

@@

A basic GLMM with spatial Gaussian random fields (GRFs) can be written as

\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\boldsymbol{\omega} &\sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
\end{aligned}
\]

where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in space $\bm{s}$ is equal to mean $\mu_s$. That mean is the result of an inverse link function $g^{-1}$ applied to a linear predictor $\eta$. In this case, that linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$ multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial GRF $\omega_{\bm{s}}$. This spatial GRF represents the effect of latent spatial variables that are not otherwise accounted for in the model. Alternatively, $\omega_{\bm{s}}$ can be thought of representing spatially correlated noise. More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected value of zero that is added to the expectation in link space.

The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from a multivariate normal distribution with covariance matrix $\boldsymbol{\Sigma}_\omega$ constrained by some covariance function that defines the rate that spatial covariance decays with
distance.
Various covariance functions are possible, but a popular and flexible function is the Mat\'ern (REF) (FIGURE TODO).
If we define $\lVert d \rVert$ as the Euclidean distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
\left(\kappa \lVert d \rVert \right)^\nu K_\nu \left( \kappa \lVert d \rVert \right).
\]
The parameter $\tau^2$ controls the spatial variance, $\Gamma$ represents the Gamma
function, $K_\nu$ represents the modified Bessel function of the second kind,
and $\kappa$ represents the spatial decorrelation rate.
The parameter $\nu$
controls the smoothness of the covariance function, but can be fixed at $\nu = 1$ to take advantage of the SPDE \citep{lindgren2011}.
The parameters $\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range as $\textrm{range} = \sqrt{8} / \kappa$ and (2) marginal standard deviation as $\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$. The range is the distance that correlation decays to $\approx 0.13$.

The assumption that correlation decays equally in all directions can be relaxed to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) = \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
\left(\kappa \lVert d \rVert \bm{H} \right)^\nu K_\nu \left( \kappa \lVert d \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated parameters governing the major axis direction of geometric anisotropy and the ratio of the major and minor axes \citep{lindgren2011, thorson2015ices, thorson2018comp}.

In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally expensive; methods of working directly with the sparse precision matrix ($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are much more efficient. We introduce the SPDE approach to working with the precision matrix in the software section, but for now, note that we can instead represent our spatial random field as
\[
\boldsymbol{\omega} \sim \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
\]
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.

%
% The SPDE approach does just that, ... have been developed lets us stay with precision and so as precision:
%
% The approach involves finding an SPDE w

@@

% Miller et al:
% The SPDE approach provides a method to approximate Q without the common substantial computational burden. The SPDE is an equation to be solved. Solutions to this equation are stochastic processes whose covariance structure is chosen to satisfy the relationship the SPDE specifies. The SPDE approach involves finding an SPDE whose solutions have the covariance structure, and implied precision matrix, that is desired for f . Lindgren et al. (2011) show how to find an approximate solution to the SPDE by representing f as a sum of basis functions multiplied by coefficients; this provides a computationally efficient way to compute Q: Lindgren et al. (2011) show that the coefficients of these basis functions form a Gaussian Markov random field, for which methods for fast computation of precision matrices already exist (Rue and Held 2005). These computations make it possible to fit models quickly using integrated nested Laplace approximations (INLA; Rue et al. 2009). Rue et al. (2017) is a comprehensive review of INLA that defines the class of latent

@@

We can extend our model to accommodate spatiotemporal data by adding Gaussian random fields for each time step $t$, $\bm{\epsilon}_{t}$

\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}_{\bm{s},t} \bm{\beta} + \omega_{\bm{s}} + \epsilon_{\bm{s},t} \right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right).
\end{aligned}
\]

In this case we assume the spatiotemporal random fields are independent at each time step, but we could alternatively assume they are structured as a random walk or autoregressive process (demonstrated in examples below). The spatiotemporal random fields represent latent variables causing spatial correlation that changes with each time step.

@@

\sloppy We then create a mesh object that contains triangulation and projection
matrices needed to apply the SPDE approach. The argument \texttt{cutoff} defines
the minimum allowed distance between mesh vertices in the units of \texttt{X}
and \texttt{Y}. \texttt{make\_mesh()} is a wrapper function for
\texttt{INLA::inla.mesh.create()} or \texttt{fmesher::fm\_rcdt\_2d\_inla()} and
\texttt{cutoff} is passed through to these functions. Alternatively, we could
have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supplied
it to the \texttt{mesh} argument in \texttt{make\_mesh()}. We recommend
constructing such a mesh for applied analyses to take advantage of additional
functionality such as coarser mesh triangles in a border around the data and
control over maximum triangle length and minimum triangle angles. We can inspect
mesh object with the associated plotting method
(Figure~\ref{fig:dog-binomial-mesh}). Our mesh has 238
(\texttt{mesh\_pcod\$mesh\$n}) vertices.

<<dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. This is a simple mesh created with the convenience function \\code{make\\_mesh()}. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
plot(mesh_pcod)
@

@@

<<pcod-fit, echo=TRUE>>=
fit_bin_rf <- sdmTMB(
  present ~ poly(log(depth), 2),
  data = pcod,
  mesh = mesh_pcod,
  spatial = "on",
  family = binomial(link = "logit")
)
@

@@

<<dog-utms, echo=TRUE>>=
dat <- add_utm_columns(dat, c("lon", "lat"),
  units = "km", utm_crs = 32609)
dat$log_depth <- log(dat$depth)
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
@

<<dog-mesh2, eval=FALSE>>=
plot(mesh)
@

>>>>

<<main-setup, include=FALSE, cache=FALSE>>=
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  autodep = TRUE,
  fig.width = 7,
  fig.asp = 0.618,
  fig.pos = "ht",
  fig.align = "center",
  cache.comments = TRUE,
  dev = "pdf",
  dpi = 140,
  optipng = "-strip all",
  highlight = FALSE
)
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
options(prompt = "R> ", continue = "+  ", width = 72, useFancyQuotes = FALSE)
opts_chunk$set(prompt = TRUE)
options(replace.assign = TRUE, width = 72, prompt = "R> ")
knitr::render_sweave()
@

<<libraries, echo=FALSE, warning=FALSE, message=FALSE>>=
library(ggplot2)
library(dplyr)
library(sdmTMB)
@

@@

A basic GLMM with spatial Gaussian random fields (GRFs) can be written as
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s}}] &= \mu_{\bm{s}},\\
\mu_{\bm{s}} &=
g^{-1} \left( \eta_{\bm{s}} \right),\\
\eta_{\bm{s}} &= \bm{X}_{\bm{s}} \bm{\beta} + \omega_{\bm{s}},\\
\boldsymbol{\omega} &\sim \operatorname{MVNormal}
  \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right),\\
\end{aligned}
\]
where the expected value $\mathbb{E}[.]$ of an observation $y$ at coordinates in
space $\bm{s}$ is equal to mean $\mu_s$. That mean is the result of an inverse
link function $g^{-1}$ applied to a linear predictor $\eta$. In this case, that
linear predictor is the combination of a model matrix $\bm{X}_{\bm{s}}$
multiplied by a vector of coefficients $\bm{\beta}$ and a value from a spatial
GRF $\omega_{\bm{s}}$. This spatial GRF represents the effect of latent spatial
variables that are not otherwise accounted for in the model. Alternatively,
$\omega_{\bm{s}}$ can be thought of representing spatially correlated noise.
More simply, $\omega_{\bm{s}}$ represents a ``wiggly'' surface with an expected
value of zero that is added to the expectation in link space.

The vector of $\omega_{\bm{s}}$ ($\bm{\omega}$) is assumed drawn from
a multivariate normal distribution with covariance matrix
$\boldsymbol{\Sigma}_\omega$ constrained by some covariance function that
defines the rate that spatial covariance decays with distance. Various
covariance functions are possible, but a popular and flexible function is the
Mat\'ern (REF) (FIGURE TODO). If we define $\lVert h \rVert$ as the Euclidean
distance between spatial points $\bm{s}_j$ and $\bm{s}_k$, we can represent the
Mat\'ern covariance $\Phi$ as
\[
\Phi\left( \bm{s}_j, \bm{s}_k \right) =
  \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \right).
\]
The parameter $\tau^2$ controls the spatial variance, $\Gamma$ represents the
Gamma function, $K_\nu$ represents the modified Bessel function of the second
kind, and $\kappa$ represents the spatial decorrelation rate. The parameter
$\nu$ controls the smoothness of the covariance function, but can be fixed at
$\nu = 1$ to take advantage of the SPDE \citep{lindgren2011}. The parameters
$\kappa$ and $\tau$ can be converted to two useful values: (1) correlation range
as $\textrm{range} = \sqrt{8} / \kappa$ and (2) marginal standard deviation as
$\sigma = 1 / \sqrt{4 \pi \exp \left(2 \log(\tau) + 2 \log(\kappa) \right) }$.
The range is the distance that correlation decays to $\approx 0.13$.

<<matern-range, warning=FALSE, message=FALSE, fig.width=9, out.width="\\textwidth", fig.asp=0.33, fig.align='center', fig.cap="TODO">>=
predictor_dat <- expand.grid(
  x = seq(0, 1, length.out = 100),
  y = seq(0, 1, length.out = 100),
  year = seq_len(6)
)
sim_mesh <- make_mesh(predictor_dat, xy_cols = c("x", "y"), cutoff = 0.01)
s1 <- sdmTMB_simulate(
  formula = ~1,
  data = predictor_dat,
  mesh = sim_mesh,
  range = 0.2,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 1,
  B = 0
)
sim_g1 <- ggplot(s1, aes(x, y, fill = mu)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  coord_fixed(expand = FALSE) +
  theme_light() +
  # theme(axis.title = element_blank()) +
  ggtitle("Range = 0.2") +
  labs(x = "X", y = "Y")

s2 <- sdmTMB_simulate(
  formula = ~1,
  data = predictor_dat,
  mesh = sim_mesh,
  range = 0.6,
  phi = 0.1,
  sigma_O = 0.2,
  seed = 1,
  B = 0
)
sim_g2 <- ggplot(s2, aes(x, y, fill = mu)) +
  geom_raster(show.legend = FALSE) +
  scale_fill_viridis_c(option = "C") +
  coord_fixed(expand = FALSE) +
  theme_light() +
  # theme(axis.title = element_blank()) +
  ggtitle("Range = 0.6") +
  labs(x = "X", y = "Y")

x <- seq(0, 1, length.out = 200)
r <- seq(0.2, 1, 0.2)
r <- c(0.2, 0.6)
df <- data.frame(
  x = rep(x, length(r)),
  range = rep(r, each = length(x))
)
matern <- function(h, sigma = 1, kappa, nu = 1) {
  ret <- (sigma^2/(2^(nu - 1) * gamma(nu))) *
    ((kappa * abs(h))^nu) *
    besselK(kappa * abs(h), nu)
  ret
  ret[x == 0] <- sigma^2
  ret
}
blues <- RColorBrewer::brewer.pal(length(r) + 1, "Blues")[-1]
df$cor <- matern(df$x, kappa = sqrt(8) / df$range)
sim_g3 <- ggplot(df, aes(x, cor, col = as.factor(range), group = as.factor(range))) +
  geom_line() +
  theme_light() +
  xlab("Distance") +
  ylab("Correlation") +
  labs(colour = "Range") +
  coord_cartesian(expand = FALSE, ylim = c(0, 1)) +
  scale_colour_manual(values = blues) +
  geom_hline(yintercept = 0.13, col = "grey50", lty = 2) +
  scale_x_continuous(breaks = r) +
  theme(legend.position = c(0.7, 0.7)) +
  ggtitle("Matérn correlation function")

cowplot::plot_grid(sim_g3, sim_g1, sim_g2, align = "h", ncol = 3L)
@

The assumption that correlation decays equally in all directions can be relaxed
to allow for geometric anisotropy
\[
\Phi\left( s_j, s_k \right) =
  \frac{\tau^2}{2^{\nu - 1}\Gamma(\nu)}
  \left(\kappa \lVert h \rVert \bm{H} \right)^\nu K_\nu
  \left( \kappa \lVert h \rVert \bm{H} \right).
\]
Here, we add a linear transformation matrix $\bm{H}$ with two estimated
parameters governing the major axis direction of geometric anisotropy and the
ratio of the major and minor axes \citep{lindgren2011, thorson2015ices,
thorson2018comp}.

In practice, inversion of the covariance matrix $\bm{\Sigma}$ is computationally
expensive and methods of working directly with the sparse precision matrix
($\bm{Q}$; $\bm{Q} = \bm{\Sigma}^{-1}$) are much more efficient. We introduce
the SPDE approach to working with the precision matrix in the software section,
but for now, note that we can instead represent our spatial random field as
\[
\boldsymbol{\omega} \sim
  \mathrm{MVNormal}\left(\boldsymbol{0}, \boldsymbol{Q}^{-1}_\omega\right),
\]
where $\boldsymbol{Q}_\omega$ is a sparse precision matrix.
%
% The SPDE approach does just that, ... have been developed lets us stay with precision and so as precision:
%
% The approach involves finding an SPDE w

@@

% Miller et al: The SPDE approach provides a method to approximate Q without the
% common substantial computational burden. The SPDE is an equation to be solved.
% Solutions to this equation are stochastic processes whose covariance structure
% is chosen to satisfy the relationship the SPDE specifies. The SPDE approach
% involves finding an SPDE whose solutions have the covariance structure, and
% implied precision matrix, that is desired for f . Lindgren et al. (2011) show
% how to find an approximate solution to the SPDE by representing f as a sum of
% basis functions multiplied by coefficients; this provides a computationally
% efficient way to compute Q: Lindgren et al. (2011) show that the coefficients
% of these basis functions form a Gaussian Markov random field, for which
% methods for fast computation of precision matrices already exist (Rue and Held
% 2005). These computations make it possible to fit models quickly using
% integrated nested Laplace approximations (INLA; Rue et al. 2009). Rue et al.
% (2017) is a comprehensive review of INLA that defines the class of latent

@@

We can extend our model to accommodate spatiotemporal data by adding Gaussian
random fields for each time step $t$, $\bm{\epsilon}_{t}$
\[
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}_{\bm{s},t} \bm{\beta} + \omega_{\bm{s}} + \epsilon_{\bm{s},t} \right),\\
\bm{\omega} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_\omega \right),\\
\bm{\epsilon}_{t} &\sim \operatorname{MVNormal} \left( \bm{0}, \bm{\Sigma}_{\epsilon} \right).
\end{aligned}
\]
In this case we assume the spatiotemporal random fields are independent at each
time step, but we could alternatively assume they are structured as a random
walk or autoregressive process (demonstrated in examples below). The
spatiotemporal random fields represent latent variables causing spatial
correlation that changes with each time step.

@@

\sloppy We then create a mesh object that contains triangulation and projection
matrices needed to apply the SPDE approach using \texttt{make\_mesh()}.
\texttt{make\_mesh()} is a wrapper function for functions in the \pkg{fmesher}
package. The argument \texttt{cutoff} defines the minimum allowed distance
between mesh vertices in the units of \texttt{X} and \texttt{Y} (km). We can
also specify additional arguments, in this case passed to
\texttt{fmesher::fm\_mesh\_2d\_inla()}: a maximum triangle edge length of 20 km
and 40 km for the inner and outer mesh, and an offset border width of 5 km and
40 km for the inner and outer meshes. Alternatively, we could have created any
mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and supply it to the
\texttt{mesh} argument in \texttt{make\_mesh()}. We can inspect mesh object with
the associated plotting method (Figure~\ref{fig:dog-binomial-mesh2}). Our mesh
has 238 (\texttt{mesh\_pcod\$mesh\$n}) vertices. Guidelines for what makes
a good mesh are described in TODO.

<<dog-binomial-mesh, results='hide', message=FALSE, warning=FALSE, echo=TRUE>>=
mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
@

<<dog-binomial-mesh2, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="SPDE mesh (lines) combined with the trawl survey observations (points). The locations where lines intersect are referred to as ``vertices''. Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. A greater degree of control over the mesh construction can be achieved by using \\pkg{fmesher} or \\proglang{R}-\\pkg{INLA} directly.", fig.width=5>>=
mesh_pcod2 <- make_mesh(
  pcod, c("X", "Y"),
  fmesher_func = fmesher::fm_mesh_2d_inla,
  cutoff = 8,
  convex = 10,
  max.edge = c(20, 40),
  offset = c(5, 40)
)
plot(mesh_pcod2)
@

@@

<<pcod-fit, echo=TRUE>>=
fit_bin_rf <- sdmTMB(
  present ~ poly(log(depth), 2),
  data = pcod,
  mesh = mesh_pcod2,
  spatial = "on",
  family = binomial(link = "logit")
)
@

@@

<<dog-utms, echo=TRUE>>=
dat <- add_utm_columns(dat, c("lon", "lat"),
  units = "km", utm_crs = 32609)
dat$log_depth <- log(dat$depth)

mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
@

<<dog-mesh2, eval=FALSE>>=
plot(mesh)
mesh$mesh$n
@
